---
title: "Effectiveness of HIV treatment combinations - Cox regression with time-varying covariates feat. `data.table`"
author: "Brandon Hao"
output: 
  md_document:
    variant: markdown_github
date: "2023-05-25"
---

```{r setup, include=FALSE}
# Set working directory
knitr::opts_chunk$set(
  root.dir = r"{C:\Users\brand\OneDrive\Desktop\Health Data Science\Datathon_team_2}"
  )

```

In this markdown, I'll demonstrate how to clean the data for a Cox regression with time-varying covarites. The data cleaning is arguably the most difficult part of the the analysis and we'll make plenty use of `data.table`'s convenient group-wise operations! I'll also demonstrate the process of fitting the Cox regression model and checking that if the proportional hazards assumption of the model is met. 

Without further ado, let's load in the necessary packages and read in the data. 

```{r initialisation, include = TRUE}
# Install and load libraries 
packages <- c('tidyverse', 'data.table', 'survival', 'DataExplorer', 'survsim', 'broom', 'survminer')

installed_packages <- packages %in% rownames(installed.packages())

if (any(installed_packages == FALSE)) {
  install.packages(packages[!installed_packages])
}

invisible(lapply(packages, require, character.only = TRUE))

# Set file paths  
if (!dir.exists('output')) dir.create(file.path('output'))

# Read HIV data
hiv <- file.path(r'{../input/HealthGymV2_CbdrhDatathon_ART4HIV.csv}') %>% fread()
```

Let's have a look at our dataset. 

```{r initial look}
# Quick look! 
glimpse(hiv)
summary(hiv)
str(hiv)
```

Firstly, the names of the columns suck! Also, there are some variables which should be categorical variables. Let's clean up the names and transform variables into their proper type. 

```{r transformation}
# Change some names coz they suck
names(hiv) <- c(
  'vl', 'cd4', 'relcd4', 'gender', 'ethnic', 'base_drug_comb', 'ini', 'nnrti', 'pi', 'pk', 'vl_m', 'cd4_m', 'drug_m', 'id', 'time'
  )

# Transform columns into Boolean
boolean_cols <- hiv %>% select(pk:drug_m) %>% names()
hiv[, (boolean_cols) := lapply(.SD, as.logical), .SDcols = boolean_cols]

# Factorise certain columns and give sensible levels
factor_cols <- hiv %>% select(gender:pi) %>% names()
hiv[, (factor_cols) := lapply(.SD, as.factor), .SDcols = factor_cols]

levels(hiv$gender) <- c('male', 'female')
levels(hiv$ethnic) <- c('asian', 'afro', 'caucasian', 'other')
levels(hiv$base_drug_comb) <- c('ftc_tdf','3tc_abc', 'ftc_taf', 'drv_ftc_tdf', 'ftc_rtvb_tdf', 'other') 
levels(hiv$ini) <- c('dtg', 'ral', 'evg', 'not_applied')
levels(hiv$nnrti) <- c('nvp', 'efv', 'rpv', 'not_applied')
levels(hiv$pi) <- c('drv', 'rtvb', 'lpv', 'rtv', 'atv', 'not_applied')

# Make sure the format is properly transformed
str(hiv)
```

Let's have a look if our participants have consistent ethnicity and sex. 

```{r ethnicity and sex, results = FALSE}
# Check if participants have same ethnicity and sex
hiv[, .(nunique_gender = uniqueN(gender), nunique_ethn = uniqueN(ethnic)), by = id][, .(max_gender = max(nunique_gender), max_ethn = max(nunique_ethn))]
```

Just do some quick automated EDA. 

```{r}
# Plotting distribution of discrete variables
hiv %>% plot_bar(by = 'ethnic', nrow = 4, title = 'EDA - Ethnicity')
hiv %>% plot_bar(by = 'gender', nrow = 4, title = 'EDA - Gender')
hiv %>% plot_bar(by = 'base_drug_comb', nrow = 4, title = 'EDA - Drug Combo')
```

Let's move forward to analysis - let's relevel some variables for later analysis.

```{r}
# Re-level some variables
hiv$gender <- relevel(hiv$gender, ref = "male")
hiv$ethnic <- relevel(hiv$ethnic, ref = "caucasian")
hiv$base_drug_comb <- relevel(hiv$base_drug_comb, ref = "ftc_tdf")
hiv$ini <- relevel(hiv$ini, ref = "not_applied")
hiv$nnrti <- relevel(hiv$nnrti, ref = "not_applied")
hiv$pi <- relevel(hiv$pi, ref = "not_applied")
```

Let's separate out each drug into it's separate binary variables.

Transform the INI, NNRTI, PI variables too. 

```{r}
# Base drug transformation
hiv_trans <- hiv %>% 
  mutate(ftc = fifelse(grepl('ftc', base_drug_comb), 1, 0),
         tdf = fifelse(grepl('tdf', base_drug_comb), 1, 0),
         tc3 = fifelse(grepl('3tc', base_drug_comb), 1, 0),
         abc = fifelse(grepl('abc', base_drug_comb), 1, 0),
         taf = fifelse(grepl('taf', base_drug_comb), 1, 0),
         drv_base = fifelse(grepl('drv', base_drug_comb), 1, 0),
         rtvb_base = fifelse(grepl('rtvb', base_drug_comb), 1, 0),
         other_base = fifelse(grepl('other', base_drug_comb), 1, 0))

# INI drug transformation
hiv_trans <- hiv_trans %>% 
  mutate(dtg = fifelse(grepl('dtg', ini), 1, 0),
         ral = fifelse(grepl('ral', ini), 1, 0),
         evg = fifelse(grepl('evg', ini), 1, 0),
         ini_not_applied = fifelse(grepl('not_applied', ini), 1, 0))

# NNTRI drug transformation
hiv_trans <- hiv_trans %>% 
  mutate(nvp = fifelse(grepl('nvp', nnrti), 1, 0),
         efv = fifelse(grepl('efv', nnrti), 1, 0),
         rpv = fifelse(grepl('rpv', nnrti), 1, 0),
         nnrti_not_applied = fifelse(grepl('not_applied', nnrti), 1, 0))

# PI drug transformation
hiv_trans <- hiv_trans %>% 
  mutate(drv_extra_pi = fifelse(grepl('drv', pi), 1, 0),
         rtvb_extra_pi = fifelse(grepl('rtvb', pi), 1, 0),
         lpv = fifelse(grepl('lpv', pi), 1, 0),
         rtv = fifelse(grepl('rtv', pi), 1, 0),
         atv = fifelse(grepl('atv', pi), 1, 0),
         pi_not_applied = fifelse(grepl('not_applied', pi), 1, 0))
```

Let's extract the NTRI regimen from the base drug combination. 

```{r}
# Extract NRTI regimen
hiv_trans <- hiv_trans %>% 
  mutate(nrti_regimen = case_when(base_drug_comb %like% '3tc' ~ '3tc + abc',
                                  base_drug_comb %like% 'taf' ~ 'ftc + taf',
                                  base_drug_comb %like% 'tdf' ~ 'ftc + tdf',
                                  .default = 'other'))
```

Let's extract the PI regimen - regardless whether it's administered as part of the base regimen or as extra PI. 

```{r}
# Extract PI regimen
pi_cols <- c("drv_base", "rtvb_base", "drv_extra_pi", "rtvb_extra_pi", "lpv", "atv", "rtv")

hiv_trans <- hiv_trans %>% 
  mutate(pi_regimen = apply(hiv_trans[, ..pi_cols], 1, function(row) {
    
    # Get column names with 1 values
    pi_columns <- names(row[row == 1])
    
    # Replace "DRV_base" with "DRV" in the column names
    pi_columns <- gsub("drv_base", "drv", pi_columns)
    
    # Replace "DRV_extra_pi" with "DRV" in the column names
    pi_columns <- gsub("drv_extra_pi", "drv", pi_columns)
    
    # Replace "RTVB_base" with "RTVB" in the column names
    pi_columns <- gsub("rtvb_base", "rtvb", pi_columns)
    
    # Replace "RTVB_extra_pi" with "RTVB" in the column names
    pi_columns <- gsub("rtvb_extra_pi", "rtvb", pi_columns)
    
    # De-duplicate pi_columns (as DRV and RTVB may appear more than once)
    pi_columns <- unique(pi_columns)
    
    # Concatenate column names with pluses in between
    regimen <- paste(pi_columns, collapse = " + ")
    
    #Return the PI regimen
    ifelse(regimen == "", "none", regimen)
  }))
```

Let's extract the INI regimen 

```{r}
# Extract INI regimen
ini_cols <- c("evg", "dtg", "ral")

hiv_trans <- hiv_trans %>% 
  mutate(ini_regimen = apply(hiv_trans[, ..ini_cols], 1, function(row) {
      
    # Get column names with 1 values
    ini_columns <- names(row[row == 1])
    
    # Concatenate column names with pluses in between
    regimen <- paste(ini_columns, collapse = " + ")
    
    # Return the INI regimen
    ifelse(regimen == "", "none", regimen)
  }))
```

Let's extract the NNRTI regimen.

```{r}
# Extract NNRTI regimen
nnrti_cols <- c("nvp", "efv", "rpv")

hiv_trans <- hiv_trans %>% 
  mutate(nnrti_regimen = apply(hiv_trans[, ..nnrti_cols], 1, function(row) {
  
    #Get column names with 1 values
    nnrti_columns <- names(row[row == 1])
    
    #Concatenate column names with pluses in between
    regimen <- paste(nnrti_columns, collapse = " + ")
    
    #Return the NRTI regimen
    ifelse(regimen == "", "none", regimen)
  }))
```

Now, we want to extract all the useful columns that we'll use for the analysis. 

```{r}
# Extract useful columns
cols_keep = c('id', 'time', 'vl', 'cd4', 'relcd4', 'gender', 'ethnic', 
              'nrti_regimen', 'pi_regimen', 'ini_regimen', 'nnrti_regimen', 'pk', 
              'vl_m', 'cd4_m', 'drug_m')

hiv_cleaned <- hiv_trans[, ..cols_keep]

# Save the processed data frame
fwrite(hiv_cleaned, file.path('output', "processed_data_hiv.csv"))
```

Factorise the non-factor columns of the cleaned dataset.

```{r}
# Factorise columns and level based on decreasing frequency
factor_cols <- hiv_cleaned %>% select(nrti_regimen:drug_m) %>% names()
hiv_cleaned[, (factor_cols) := lapply(.SD, factor), .SDcols = factor_cols]
hiv_cleaned[, (factor_cols) := lapply(.SD, fct_infreq), .SDcols = factor_cols]

str(hiv_cleaned)
```

# Start analysis proper

Let's first look at how many people changed their regimen? 

```{r}
# How many patients change their regimen at least once?
npatient_regimen_change <- hiv_cleaned[, .N, by = .(id, nrti_regimen, pi_regimen, ini_regimen, nnrti_regimen, pk)][, .(n_regimens = .N), by = id][n_regimens > 1, .N]

# Total number of patients
npatients <- hiv_cleaned[, uniqueN(id)]

c('total unique patients' = npatients, 'patients with regimen changes' = npatient_regimen_change)
```

Out of `r npatients`, `r npatient_regimen_change` changed regiments, representing `r npatient_regimen_change / (npatients) * 100 %>% round(2)`% of the cohort

As we are investigating both viral load and immune recovery outcomes, split dataset to 1) people over 1000 VL at time zero and 2) people below 500 CD4 at time zero.

```{r}
# Get our VL dataset by getting unique patients who started with >1000 VL at time 0
vl_id <- hiv_cleaned[vl > 1000 & time == 0, unique(id)]
vl <- hiv_cleaned[id %in% vl_id]

# Get our CD4 dataset by getting unique patients who started with <500 CD4 at time 0
cd4_id <- hiv_cleaned[cd4 < 500 & time == 0, unique(id)]
cd4 <- hiv_cleaned[id %in% cd4_id]
```

Now we want to restrict the dataset so the dataset only contains rows for each patient up to their censor/first_event timepoint. 

```{r}
# Give outcomes and mark timing for when the event occurred
vl[, `:=`(event = vl <= 1000, event_timing = ifelse(vl <= 1000, time, 59))]
cd4[, `:=`(event = cd4 >= 500, event_timing = ifelse(cd4 >= 500, time, 59))]

# Now for each patient, grab their first event timing
vl_first_event <- vl[, .(first_event = min(event_timing)), by = id]
cd4_first_event <- cd4[, .(first_event = min(event_timing)), by = id]

# Create trimmed dataset
vl_trim <- vl %>% 
  left_join(vl_first_event, by = 'id') %>% 
  filter(time <= first_event) %>% 
  select(-event_timing)

cd4_trim <- cd4 %>% 
  left_join(cd4_first_event, by = 'id') %>% 
  filter(time <= first_event) %>% 
  select(-event_timing)
```

Create an input dataset for modelling involving start-end date for each patient and their treatment regimen. For details, see comments in code below.

```{r}
# Get distinct episodes for each patient
# `episode_cumu` column only stays the same if there are no patient changes nor treatment regimen changes  
vl_episode <- vl_trim %>% 
  mutate(episode = fifelse(id == lag(id, default = first(id)) 
                           & nrti_regimen == lag(nrti_regimen, default = first(nrti_regimen))
                           & pi_regimen == lag(pi_regimen, default = first(pi_regimen))
                           & ini_regimen == lag(ini_regimen, default = first(ini_regimen))
                           & nnrti_regimen == lag(nnrti_regimen, default = first(nnrti_regimen))
                           & pk == lag(pk, default = first(pk)),
                           0, 1),
         episode_cumu = cumsum(episode)
         )

# Getting start and stop times for each independent episode
splits <- vl_episode[, .(start = min(time), time = max(time)), by = .(id, episode_cumu)]

# Inner join so that everyone only has rows indicating the start and end timepoint of each distinct episode
vl_final <- vl_episode %>% 
  inner_join(splits, by = c('id', 'time', 'episode_cumu')) %>% 
  mutate(stop = time) %>% 
  select(-episode, -episode_cumu)

# Since current censoring timepoint is the timepoint immediate prior to change, 
# we want to increment current stop timepoint by 1 to indicate the actual censoring timepoint 
vl_final[, stop := stop + 1]

# Of course, we don't want to increment the ultimate censoring timepoint at which
# people reached the event or past timepoint 59.
# Thus we limit max stop point to the timepoint of ultimate patient censoring 
vl_final[, stop := fifelse(stop > first_event, first_event, stop)]

# Re-create event flag to align against new stop date
# See patient ID 2 as example
# Patient 2 reached <1000 VL at time 37
# However at the same timepoint they also changed PI regimens
# Now, does their regimen change count as a line?
# It shouldn't since we're assuming that he reached <1000 VL at timepoint 37
# before he changed medication 
# Thus we'd keep his first regimen but discard his second
# This is the purpose of the code below for creating new flag then dedupe
vl_final <- vl_final[, event := fifelse(stop == first_event, TRUE, FALSE)] %>% 
  distinct(id, event, .keep_all = TRUE)
```

Repeat the same process for CD4 model. 

```{r}
# For more details see code above 
cd4_episode <- cd4_trim %>% 
  mutate(episode = fifelse(id == lag(id, default = first(id)) 
                           & nrti_regimen == lag(nrti_regimen, default = first(nrti_regimen))
                           & pi_regimen == lag(pi_regimen, default = first(pi_regimen))
                           & ini_regimen == lag(ini_regimen, default = first(ini_regimen))
                           & nnrti_regimen == lag(nnrti_regimen, default = first(nnrti_regimen))
                           & pk == lag(pk, default = first(pk)),
                           0, 1),
         episode_cumu = cumsum(episode)
         )


splits <- cd4_episode[, .(start = min(time), time = max(time)), by = .(id, episode_cumu)]


cd4_final <- cd4_episode %>% 
  inner_join(splits, by = c('id', 'time', 'episode_cumu')) %>% 
  mutate(stop = time) %>% 
  select(-episode, -episode_cumu)

cd4_final[, stop := stop + 1]

cd4_final[, stop := fifelse(stop > first_event, first_event, stop)]

cd4_final <- cd4_final[, event := fifelse(stop == first_event, TRUE, FALSE)] %>% 
  distinct(id, event, .keep_all = TRUE)
```

# Build Cox model!!!!!!! 

Let's build time-varying Cox model starting with Viral load model.

```{r}
# Build time-varying Cox model for VL
vl_mod <- coxph(Surv(start, stop, event) ~ relcd4 + gender + ethnic + nrti_regimen + pi_regimen + ini_regimen + nnrti_regimen + pk + cluster(id), data = vl_final)

# Display results
results <- tidy(vl_mod, conf.int = TRUE, exp = T) 
results %>% fwrite(file.path('output', 'vl_results.csv'))
summary(vl_mod)

sjPlot::plot_model(vl_mod)
ggsave(file.path('output', 'vl_results_image.png'))

results

vl_mod
```

It's seems like the cox proportional hazards assumption does not hold for the VL-model:

```{r}

ggcoxzph((cox.zph(vl_mod)))[1]
ggcoxzph((cox.zph(vl_mod)))[2]
ggcoxzph((cox.zph(vl_mod)))[3]
ggcoxzph((cox.zph(vl_mod)))[4]
ggcoxzph((cox.zph(vl_mod)))[5]
ggcoxzph((cox.zph(vl_mod)))[6]
ggcoxzph((cox.zph(vl_mod)))[7]
ggcoxzph((cox.zph(vl_mod)))[8]

```

Let's build time-varying Cox model starting with CD4 model.

```{r}
# Build time-varying Cox model for CD4
cd4_mod <- coxph(Surv(start, stop, event) ~ vl + gender + ethnic + nrti_regimen + pi_regimen + ini_regimen + nnrti_regimen + pk + cluster(id), data = cd4_final)

# Display results
results <- tidy(cd4_mod, conf.int = TRUE, exp = T) 
results %>% fwrite(file.path('output', 'cd4_results.csv'))
summary(cd4_mod)

sjPlot::plot_model(cd4_mod)
ggsave(file.path('output', 'cd4_results_image.png'))

results
```


It's seems like the cox proportional hazards assumption does not hold also for the CD4-model:

```{r}

ggcoxzph((cox.zph(cd4_mod)))[1]
ggcoxzph((cox.zph(cd4_mod)))[2]
ggcoxzph((cox.zph(cd4_mod)))[3]
ggcoxzph((cox.zph(cd4_mod)))[4]
ggcoxzph((cox.zph(cd4_mod)))[5]
ggcoxzph((cox.zph(cd4_mod)))[6]
ggcoxzph((cox.zph(cd4_mod)))[7]
ggcoxzph((cox.zph(cd4_mod)))[8]

```

* The only variable for which the cox proportional hazard assumption seems to hold is the pk-enhancer variable in both models (although the Schoenfeld test is barely non-significant in the CD4-model).
