---
title: "HIV_analysis"
output: html_document
date: "2023-05-25"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Let's start by initialising our analysis. Load some libraries, define some file paths and read the data.

```{r initialisation, results = FALSE}

# Set working directory
base_folder <- ("C:/Users/Brandon/OneDrive - The University of Melbourne/Desktop/Health Data Science/Datathon/DATATHON_TEAM_2")

# Install and load libraries 
packages <- c('tidyverse', 'data.table', 'survial', 'DataExplorer')
installed_packages <- packages %in% rownames(installed.packages())

if (any(installed_packages == FALSE)) {
  install.packages(packages[!installed_packages])
}

invisible(lapply(packages, require, character.only = TRUE))

# Set file paths  
output_path <- file.path(base_folder, 'output')

if (!dir.exists(output_path)) {
  dir.create(file.path(output_path))
}

# Read HIV data
hiv <- fread(file.path(base_folder, 'input', 'HealthGymV2_CbdrhDatathon_ART4HIV.csv'))
```
Let's have a look at our dataset. 

```{r initial look}

# Quick look! 
glimpse(hiv)
summary(hiv)
str(hiv)
```

Firstly, the names of the columns suck! Also, there are some variables which should be categorical variables. Let's clean up the names and transform some variables into their proper type. 

```{r transformation}

# Change some names coz they suck
names(hiv) <- c('vl', 'cd4', 'relcd4', 'gender', 'ethnic', 'base_drug_comb', 'ini', 'nnrti', 'pi', 'pk', 'vl_m', 'cd4_m', 'drug_m', 'id', 'time')

# Change patient ID to start with 1 to be more intuitive
hiv$id <- hiv$id + 1

# Transform columns into Boolean
boolean_cols <- hiv %>% select(pk:drug_m) %>% names()
hiv[, (boolean_cols) := lapply(.SD, as.logical), .SDcols = boolean_cols]

# Factorise certain columns and give sensible levels
factor_cols <- hiv %>% select(gender:pi) %>% names()
hiv[, (factor_cols) := lapply(.SD, as.factor), .SDcols = factor_cols]

levels(hiv$gender) <- c('male', 'female')
levels(hiv$ethnic) <- c('asian', 'afro', 'caucasian', 'other')
levels(hiv$base_drug_comb) <- c('ftc_tdf','3tc_abc', 'ftc_taf', 'drv_ftc_tdf', 'ftc_rtvb_tdf', 'other') 
levels(hiv$ini) <- c('dtg', 'ral', 'evg', 'not_applied')
levels(hiv$nnrti) <- c('nvp', 'efv', 'rpv', 'not_applied')
levels(hiv$pi) <- c('drv', 'rtvb', 'lpv', 'rtv', 'atv', 'not_applied')

# Make sure the format is properly transformed
str(hiv)
```

Let's have a look if our participants have consistent ethnicity and sex. 

```{r ethnicity and sex, results = FALSE}
# Check if participants have same ethnicity and sex
hiv[, .(nunique_gender = uniqueN(gender), nunique_ethn = uniqueN(ethnic)), by = id]
```

Just do some quick automated EDA. 

```{r}
# Automated EDA
# hiv %>% create_report(
#   #output_file = paste("Report", format(Sys.time(), "%d-%m-%Y %H:%M:%S %Z"), sep = ' - '),
#   report_title = 'EDA Report - HIV Dataset',
#   output_dir = output_path
# )

# Plotting distribution of discrete variables
hiv %>% plot_bar(by = 'ethnic', nrow = 4, title = 'EDA - Ethnicity')
hiv %>% plot_bar(by = 'gender', nrow = 4, title = 'EDA - Gender')
hiv %>% plot_bar(by = 'base_drug_comb', nrow = 4, title = 'EDA - Drug Combo')
```

Let's move forward to analysis - let's relevel some variables for later analysis.

```{r}
#Re-level some variables
hiv$gender <- relevel(hiv$gender, ref = "male")
hiv$ethnic <- relevel(hiv$ethnic, ref = "other")
hiv$base_drug_comb <- relevel(hiv$base_drug_comb, ref = "other")
hiv$ini <- relevel(hiv$ini, ref = "not_applied")
hiv$nnrti <- relevel(hiv$nnrti, ref = "not_applied")
hiv$pi <- relevel(hiv$pi, ref = "not_applied")
```

Let's separate out each drug into it's separate binary variables.

Transform the INI, NNRTI, PI variables too. 

```{r}
# Base drug transformation
hiv_trans <- hiv %>% 
  mutate(ftc = fifelse(grepl('ftc', base_drug_comb), 1, 0),
         tdf = fifelse(grepl('tdf', base_drug_comb), 1, 0),
         tc3 = fifelse(grepl('3tc', base_drug_comb), 1, 0),
         abc = fifelse(grepl('abc', base_drug_comb), 1, 0),
         taf = fifelse(grepl('taf', base_drug_comb), 1, 0),
         drv_base = fifelse(grepl('drv', base_drug_comb), 1, 0),
         rtvb_base = fifelse(grepl('rtvb', base_drug_comb), 1, 0),
         other_base = fifelse(grepl('other', base_drug_comb), 1, 0))

# INI drug transformation
hiv_trans <- hiv_trans %>% 
  mutate(dtg = fifelse(grepl('dtg', ini), 1, 0),
         ral = fifelse(grepl('ral', ini), 1, 0),
         evg = fifelse(grepl('evg', ini), 1, 0),
         ini_not_applied = fifelse(grepl('not_applied', ini), 1, 0))

# NNTRI drug transformation
hiv_trans <- hiv_trans %>% 
  mutate(nvp = fifelse(grepl('nvp', nnrti), 1, 0),
         efv = fifelse(grepl('efv', nnrti), 1, 0),
         rpv = fifelse(grepl('rpv', nnrti), 1, 0),
         nnrti_not_applied = fifelse(grepl('not_applied', nnrti), 1, 0))

# PI drug transformation
hiv_trans <- hiv_trans %>% 
  mutate(drv_extra_pi = fifelse(grepl('drv', pi), 1, 0),
         rtvb_extra_pi = fifelse(grepl('rtvb', pi), 1, 0),
         lpv = fifelse(grepl('lpv', pi), 1, 0),
         rtv = fifelse(grepl('rtv', pi), 1, 0),
         atv = fifelse(grepl('atv', pi), 1, 0),
         pi_not_applied = fifelse(grepl('not_applied', pi), 1, 0))
```

Now, concatenate the drugs row-wise 

```{r}
# Concatenate string row-wise
hiv_trans$nrti_regimen <- apply(hiv_trans[, c("ftc", "tdf", "tc3", "abc", "taf")], 1, function(row) {
  
  #Get column names with 1 values
  nrti_columns <- names(row[row == 1])
  
  #Replace "TC3" with "3TC" in the column names
  nrti_columns <- gsub("tc3", "3tc", nrti_columns)
  
  #Concatenate column names with pluses in between
  regimen <- paste(nrti_columns, collapse = " + ")
  
  #Return the NRTI regimen
  ifelse(regimen == "", "none", regimen)
})
```


```{r}
#Create a new variable `PI_Regimen` based on specific columns
hiv_trans$pi_regimen <- apply(hiv_trans[, c("drv_base", "rtvb_base", "drv_extra_pi", "rtvb_extra_pi", "lpv", "atv", "rtv")], 1, function(row) {
  
  #Get column names with 1 values
  pi_columns <- names(row[row == 1])
  
  #Replace "DRV_base" with "DRV" in the column names
  pi_columns <- gsub("drv_base", "drv", pi_columns)
  
  #Replace "DRV_extra_pi" with "DRV" in the column names
  pi_columns <- gsub("drv_extra_pi", "drv", pi_columns)
  
  #Replace "RTVB_base" with "RTVB" in the column names
  pi_columns <- gsub("rtvb_base", "rtvb", pi_columns)
  
  #Replace "RTVB_extra_pi" with "RTVB" in the column names
  pi_columns <- gsub("rtvb_extra_pi", "rtvb", pi_columns)
  
  #De-duplicate pi_columns (as DRV and RTVB may appear more than once)
  pi_columns <- unique(pi_columns)
  
  #Concatenate column names with pluses in between
  regimen <- paste(pi_columns, collapse = " + ")
  
  #Return the NRTI regimen
  ifelse(regimen == "", "none", regimen)
})
```

```{r}
#Create a new variable `INSTI_regimen` based on specific columns
hiv_trans$ini_regimen <- apply(hiv_trans[, c("evg", "dtg", "ral")], 1, function(row) {
  
  #Get column names with 1 values
  insti_columns <- names(row[row == 1])
  
  #Concatenate column names with pluses in between
  regimen <- paste(insti_columns, collapse = " + ")
  
  #Return the NRTI regimen
  ifelse(regimen == "", "none", regimen)
  
})
```

```{r}
#Create a new variable `NNRTI_regimen` based on specific columns
hiv_trans$nnrti_regimen <- apply(hiv_trans[, c("nvp", "efv", "rpv")], 1, function(row) {
  
  #Get column names with 1 values
  nnrti_columns <- names(row[row == 1])
  
  #Concatenate column names with pluses in between
  regimen <- paste(nnrti_columns, collapse = " + ")
  
  #Return the NRTI regimen
  ifelse(regimen == "", "none", regimen)
  
})
```

```{r}
hiv_cleaned <- hiv_trans %>% as.data.frame()

#Change the order of the columns
other_columns <- setdiff(names(hiv_cleaned), c("pk", "id", "time"))

#Rearrange the columns
hiv_cleaned[c(other_columns, c("pk", "id", "time"))] -> hiv_cleaned
  
#Save the processed data frame
fwrite(hiv_cleaned, file.path(output_path, "processed_data_hiv.csv"))
```

Factorise the cleaned dataset.

```{r}
# Subset the columns we'd like to keep
cols_keep = c('vl', 'cd4', 'relcd4', 'gender', 'ethnic', 
              'nrti_regimen', 'other_base', 'pi_regimen', 'ini_regimen', 'nnrti_regimen', 'pk',
              'id', 'time', 'vl_m', 'cd4_m', 'drug_m')

hiv_trimmed <- hiv_cleaned[, cols_keep] %>% as.data.table()

# Factorise certain columns and give sensible levels
factor_cols <- hiv_trimmed %>% select(nrti_regimen:pk) %>% names()
hiv_trimmed[, (factor_cols) := lapply(.SD, as.factor), .SDcols = factor_cols]

str(hiv_trimmed)
```

```{r}
# Refactor variable
hiv_trimmed$nrti_regimen <- fct_relevel(
  hiv_trimmed$nrti_regimen, hiv_trimmed$nrti_regimen %>% table() %>% sort(decreasing = TRUE) %>% names()
  )

hiv_trimmed$other_base <- fct_relevel(
  hiv_trimmed$other_base, hiv_trimmed$other_base %>% table() %>% sort(decreasing = TRUE) %>% names()
  )

hiv_trimmed$pi_regimen <- fct_relevel(
  hiv_trimmed$pi_regimen, hiv_trimmed$pi_regimen %>% table() %>% sort(decreasing = TRUE) %>% names()
  )

hiv_trimmed$ini_regimen <- fct_relevel(
  hiv_trimmed$ini_regimen, hiv_trimmed$ini_regimen %>% table() %>% sort(decreasing = TRUE) %>% names()
  )

hiv_trimmed$nnrti_regimen <- fct_relevel(
  hiv_trimmed$nnrti_regimen, hiv_trimmed$nnrti_regimen %>% table() %>% sort(decreasing = TRUE) %>% names()
  )

hiv_trimmed$pk <- fct_relevel(
  hiv_trimmed$pk, hiv_trimmed$pk %>% table() %>% sort(decreasing = TRUE) %>% names()
  )

str(hiv_trimmed)
```

# Start analysis proper

How many people change their reigmen? 

```{r}
# How many patients change their regimen at least once?
npatient_regimen_change <- hiv_trimmed[, .N, by = .(id, nrti_regimen, other_base, pi_regimen, ini_regimen, nnrti_regimen, pk)] %>% 
  group_by(id) %>% 
  summarise(n_regimens = n()) %>% 
  filter(n_regimens > 1) %>% 
  nrow()

npatients <- uniqueN(hiv_trimmed$id)

npatients; npatient_regimen_change
```
Out of `r npatients`, `r npatient_regimen_change` changed regiments, representing `r npatient_regimen_change / (npatients) * 100 %>% round(2)`% of the cohort

Split dataset to 1) people over 1000 VL at time zero and 2) people below 500 CD4 at time zero


```{r}
# Get our VL dataset 
vl_id <- hiv_trimmed %>% filter(vl > 1000 & time == 0) %>% select(id) %>% unique()
vl <- hiv_trimmed %>% filter(id %in% vl_id$id)

# Give outcome
vl <- vl %>% mutate(event = vl < 1000)
```

Now, transform to format for survival analysis

```{r}
# Give out
```







```{r}
treatmemt_comb <- c('nrti_regimen', 'other_base', 'pi_regimen', 'ini_regimen', 'nnrti_regimen', 'pk')

hiv_trimmed %>% 
  group_by(id, nrti_regimen, other_base, pi_regimen, ini_regimen, nnrti_regimen, pk) %>% 
  summarise(start = first(time), end = last(time))
```
```{r}
hiv_trimmed %>% filter(id == 6)
```








